### 2022-07-24 paper/localpsf.tex
more writing in numerical results section

\begin{figure}[H]
	\begin{subfigure}{0.5\textwidth}
		\includegraphics[scale=0.23]{meshHeight_view2_edges.png}
		\caption{Bird's eye view}
	\end{subfigure}
	\begin{subfigure}{0.48\textwidth}
		\includegraphics[scale=0.22]{meshHeight_view0_edges_withAxes.png}
		\caption{Top-down view}
	\end{subfigure}
	\caption{The ice sheet, discretized by a mesh of tetrahedra. Color indicates the height of the base of the ice sheet (i.e., the mountain topography).}
	\label{fig:ice_mountain_mesh}
\end{figure}

@@

The true parameter field, $\basalfriction_\text{true}$, and corresponding velocity field, $\bm{u}_\text{true}$, are shown in Figure \ref{fig:true_beta_u}. We generate synthetic observations by restricting the tangential component of $\bm{u}_\text{true}$ to the top surface to get
\begin{equation*}
\bm{y}_\text{true} = \bm{T} \bm{u}_\text{true} |_{\Gamma_t}.
\end{equation*}
We add $1\%$ multiplicative gaussian noise, $\boldsymbol{\xi}$, to all components of the finite element vector representing $\bm{y}_\text{true}$ to get noisy synthetic observations, $\bm{y}_\text{obs}$, which are used for the inversion. That is, 
\begin{equation*}
\left[\bm{y}_\text{obs}\right]_j := \left[\bm{y}_\text{true}\right]_j + \epsilon \left[\boldsymbol{\xi}\right]_j \left|\left[\bm{y}_\text{true}\right]_j\right|
\end{equation*}
where $\epsilon=0.01$, and the components of the noise function, $\boldsymbol{\xi}_j \sim N(0,1)$, are independent and identically distributed (i.i.d) random numbers drawn from the standard normal distribution. Here $\left[\bm{f}\right]_j$ denotes the $j$th entries of the finite element vector for the discretized versions of the function $\bm{f}:\Gamma_t \rightarrow \mathbb{R}^2$, and $j$ ranges over all entries of the vector. Here we do not use bold to distinguish finite element vectors from their associated functions (as we did previously) because the functions $\bm{y}_\text{true}$, $\bm{y}_\text{obs}$, and $\bm{\xi}$ are already written in bold to denote the fact that they are vector valued.

The inverse problem that we seek to solve takes the form
\begin{equation}
	\min_{\basalfriction} \quad J(\basalfriction) = J_d(\basalfriction) + J_r(\basalfriction).
\end{equation}
The first term,
\begin{equation*}
	J_d(\basalfriction) := \frac{1}{2}\int_{\Gamma_{t}}||\bm{d} - \bm{T}\bm{u}||_{2}^{2}\,dS,
\end{equation*}
is the data misfit, which measures the mismatch between the observed data and the predicted data based on a candidate parameter $\basalfriction$. Here $\bm{u} = \bm{u}(\basalfriction)$ denotes the velocity field solving $\ref{Stokeseqn}$ for the given parameter $\basalfriction$. The second term,
\begin{equation*}
	J_r(\basalfriction) := \frac{1}{2}\int_{\Omega}\basalfriction\mathcal{K}^{2}\basalfriction\,dx
\end{equation*}
is a Bilaplacian regularization term. Specifically, $\mathcal{K}$ is the inverse of the solution operator for the following elliptic PDE:
\begin{align*}
-\gamma\Delta u +\delta\,u&=s\,\,\,\,\text{ in }\Gamma_{b},\\
\gamma \bm{\nabla}u\cdot\bm{n}+\mu\,u&=0\,\,\,\text{ on }\partial\Gamma_{b}.
\end{align*}
The constant $\delta$ is chosen so that the correlation length of functions drawn from $N(0,\mathcal{L}^{-2})$ is $1/10$th of the radius of the domain. In all results shown except one, $\gamma$ is chosen so that the solution satisfies the Morozov discrepancy principle, that is,
\begin{equation*}
	\frac{1}{2}\int_{\Gamma_{t}}||\bm{d} - \bm{T}\bm{u}||_{2}^{2}\,dS = \frac{1}{2}\int_{\Gamma_{t}}||\boldsymbol{\xi}||_{2}^{2}\,dS.
\end{equation*}
The exception is Figure \ref{fig:gamma_sweep}, where we vary $\gamma$ to study the effect of changing $\gamma$ on the quality of our preconditioner.

The inverse problem is solved with an inexact-Newton conjugate gradient (CG) scheme. The initial guess for the parameter, $\basalfriction_0$, is the constant function with value $\basalfriction_0(x)=10.5$, which is halfway between the maximum and minimum values for $\basalfriction_\text{true}$. The parameter value is updated iteratively according to the formula
\begin{equation*}
	\basalfriction_{k+1} = \basalfriction_k + \alpha_k p_k,
\end{equation*}
where the search direction $p_k$ is an approximate solution to the Newton linear system
\begin{equation}
	\label{eq:newton_system}
	H p_k = - g_k,
\end{equation}
and $\alpha_k \in \mathbb{R}$ is a step length parameter chosen to satisfy the Armijo conditions \cite{ARMIJO}. Here $g_k$ is the gradient of $J$ with respect to $\basalfriction$ at $\basalfriction_k$, and $H$ is either the Hessian or Gauss-Newton Hessian of $J$ with respect to $\basalfriction$ at $\basalfriction_k$. The Gauss-Newton Hessian is used for the first five iterations, and the Hessian is used for all subsequent iterations. The method terminates when the norm of the gradient drops by a factor of $10^8$. The Newton system, \eqref{eq:newton_system}, is solved inexactly to the tolerance $\min \left(0.5, \sqrt{||g_k||}\right)||g_k||$, using the preconditioned conjugate gradient method, which only requires the application of the (Gauss-Newton) Hessian to vectors. Each application of the (Gauss-Newton) Hessian to a vector requires solving one forward and one adjoint Stokes PDE of the form \eqref{Stokeseqn} , but with different right hand sides.

We use the framework described in this paper to generate Hessian preconditioners. We use our methodology to build H-matrix approximations of the data-misfit component of the Gauss-Newton Hessian. That is, the term in the Gauss-Newton Hessian that arises from the data misfit term $J_d$. We approximate the Gauss-Newton Hessian rather than the Hessian because the Hessian often has negative values in its integral kernel when one is away from the optimal point. At the discrete level, straightforward calculation shows that the regularization term in both the Hessian and the Gauss-Newton Hessian is given by
\begin{equation*}
	\bm{K} \bm{M}^{-1} \bm{K},
\end{equation*}
where $\bm{K}$ is the finite element discretization of the elliptic operator $\mathcal{K}$, and $\bm{M}$ is the finite element mass matrix. These matrices are sparse, so we form H-matrix approximations of these matrices using standard techniques for approximating sparse matrices with H-matrices, then invert $\bm{M}^{-1}$, and multiply these matrices together using H-matrix methods. We then add this H-matrix to the H-matrix for the data misfit term constructed using our method, and factor the resulting Hessian approximation using fast H-matrix factorization methods \cite{HLIBPRO}. We construct preconditioners using our method with 1, 5, and 25 impulse responses batches. The resulting preconditioners are denoted by PSF (1), PSF (5), and PSF (25), respectively. We compare the performance of these preconditioners with the most commonly used existing preconditioners, which are no preconditioning (denoted by NONE), and preconditioning by the regularization operator (denoted REG).

\begin{figure}[H]
	\begin{subfigure}{0.24\textwidth}
		\includegraphics[scale=0.1]{m1_psf.png}
		%		\caption{}
		%		\label{fig:newtoniter1}
	\end{subfigure}
	\begin{subfigure}{0.24\textwidth}
		\includegraphics[scale=0.1]{m3_psf.png}
		%		\caption{}
		%		\label{fig:newtoniter2}
	\end{subfigure}
	\begin{subfigure}{0.24\textwidth}
		\includegraphics[scale=0.1]{m4_psf.png}
		%		\caption{}
		%		\label{fig:newtoniter3}
	\end{subfigure}
	\begin{subfigure}{0.24\textwidth}
		\includegraphics[scale=0.1]{m6_psf.png}
		%		\caption{}
		%		\label{fig:newtoniter4}
	\end{subfigure}
	%\begin{subfigure}{0.05\textwidth}
	%	\includegraphics[scale=0.1]{mColorBar.png}
	%\end{subfigure}
	\caption{The log basal friction parameter $\basalfriction$ after $1$ (left), $3$ (second from left), and $4$ (third from left) Newton iterates and the final Newton iterate (right). The preconditioner is constructed between Newton iterations 3 and 4.} 
\end{figure}

\begin{table}[H]
	\begin{center}
		\begingroup
		\setlength{\tabcolsep}{3pt}
		\renewcommand{\arraystretch}{1.25}
		\begin{tabular}{c| c c c | c c c | c c c}
			&  \multicolumn{3}{|c|}{PSF (5)} & \multicolumn{3}{|c|}{REG} & \multicolumn{3}{|c}{NONE} \\
			\hline
			Iter & 
			\#CG & \#Stokes & $||g||$ & 
			\#CG & \#Stokes & $||g||$ & 
			\#CG & \#Stokes & $||g||$ \\
			\verb~0~ &
			\verb~1~ & \verb~4~ & \verb~2.0e+7~ &
			\verb~1~ & \verb~4~ & \verb~2.0e+7~ &
			\verb~1~ & \verb~4~ & \verb~2.0e+7~ \\
			\verb~1~ &
			\verb~2~ & \verb~6~ & \verb~5.8e+6~ &
			\verb~2~ & \verb~6~ & \verb~7.9e+6~ &
			\verb~2~ & \verb~6~ & \verb~5.8e+6~ \\
			\verb~2~ &
			\verb~4~ & \verb~10~ & \verb~2.4e+6~ &
			\verb~5~ & \verb~12~ & \verb~3.9e+6~ &
			\verb~4~ & \verb~10~ & \verb~2.4e+6~ \\
			\verb~3~ &
			\verb~2~ & \verb~6+22~ & \verb~5.8e+5~ &
			\verb~13~ & \verb~28~ & \verb~1.6e+6~ &
			\verb~13~ & \verb~28~ & \verb~5.8e+5~ \\
			\verb~4~ &
			\verb~5~ & \verb~12~ & \verb~5.6e+4~ &
			\verb~42~ & \verb~86~ & \verb~4.8e+5~ &
			\verb~33~ & \verb~68~ & \verb~9.1e+4~ \\
			\verb~5~ &
			\verb~10~ & \verb~22~ & \verb~3.5e+3~ &
			\verb~84~ & \verb~170~ & \verb~7.7e+4~ &
			\verb~57~ & \verb~116~ & \verb~6.2e+3~ \\
			\verb~6~ &
			\verb~14~ & \verb~30~ & \verb~2.7e+1~ &
			\verb~125~ & \verb~252~ & \verb~5.0e+3~ &
			\verb~76~ & \verb~154~ & \verb~1.1e+2~ \\
			\verb~7~ &
			\verb~0~ & \verb~2~ & \verb~4.1e-2~ &
			\verb~194~ & \verb~390~ & \verb~7.9e+1~ &
			\verb~117~ & \verb~236~ & \verb~2.3e-1~ \\
			\verb~8~ & 
			\verb~---~ & \verb~---~ & \verb~---~ &
			\verb~0~ & \verb~2~ & \verb~1.6e-1~ &
			\verb~0~ & \verb~2~ & \verb~4.1e-2~ \\
			\hline
			Total & 
			\verb~38~ & \verb~114~ & \verb~---~ &
			\verb~466~ & \verb~950~ & \verb~---~ &
			\verb~303~ & \verb~624~ & \verb~---~ \\
		\end{tabular}
		\endgroup
	\end{center}
	\caption{Convergence history for solving the Stokes inverse problem using the preconditioned inexact Newton-Krylov method. The Newton iteration terminates when the norm of the gradient has decreased from its initial value by a factor of $10^{-8}$. The Gauss-Newton Hessian is used in iterations $0$--$5$, and the Hessian is used in all subsequent iterations. Preconditioners shown are our method with five impulse response batches (PSF (5)), regularization preconditioning (REG), and no preconditioning (NONE). Our preconditioner (PSF (5)) is built at iteration $3$ using the Gauss-Newton Hessian, and re-used for all subsequent iterations. The iterates are $\basalfriction_0$ (initial guess), $\basalfriction_1$ (parameter after one Newton step), $\basalfriction_2$ (parameter after two Newton steps), and so on. The columns titled \#CG show the number of conjugate gradient iterations used to solve the Newton system at $\basalfriction_k$, for the purpose of computing a search direction towards $\basalfriction_{k+1}$. The columns titled $||g||$ show the $l^2$ norm of the gradient at $\basalfriction_k$. The columns titled \#Stokes show the number of Stokes PDE solves performed in each Newton iteration. This consists of Stokes solves for $u$ performed during the linesearch going from $\basalfriction_{k-1}$ to $\basalfriction_k$ (including one Stokes solve at $\basalfriction_k$), plus one adjoint Stokes solve to compute the gradient at $\basalfriction_k$, plus one incremental forward and one incremental adjoint Stokes solve per CG iteration for solving the Newton system at iteration $k$. In the PSF (5) portion of row 3, we write $6+22$ to indicate that 6 Stokes solves were used during the standard course of the iteration, and 22 additional Stokes solves were used to build our preconditioner. }
\end{table}

\begin{figure}[H]
	\begin{subfigure}{0.49\textwidth}
		\centering
		\begin{tikzpicture}[baseline, trim axis right]
		\begin{axis}[width=0.99\linewidth, height=1.0\linewidth,
		ymode=log,
		ymin=1e-11,
		ymax=2,
		xmax=750,
		axis x line*=bottom,
		axis y line*=left,
		xtick={0, 200, 400, 600},
		xlabel={CG iteration},
		ylabel={Relative residual norm},
		title={CG convergence},
		legend style={at={(1.1,1)},anchor=north east},
		%legend pos=north east,
		legend style={font=\small}
		]
		\addlegendentry{PSF (1)}
		\addplot[blue, line width=0.5mm] table {cg_relres_p1.dat};
		\addlegendentry{PSF (5)}
		\addplot[red, line width=0.5mm] table {cg_relres_p5.dat};
		\addlegendentry{PSF (25)}
		\addplot[purple, line width=0.5mm] table {cg_relres_p25.dat};
		\addlegendentry{Reg}
		\addplot[dotted, line width=0.5mm] table {cg_relres_reg.dat};
		\addlegendentry{None}
		\addplot[dash dot, line width=0.5mm] table {cg_relres_none.dat};
		\end{axis}
		\end{tikzpicture}
		\caption{}
		\label{fig:cg_convergence}
	\end{subfigure}
	\begin{subfigure}{0.49\textwidth}
		\centering
		\begin{tikzpicture}[baseline, trim axis right]
		\begin{axis}[width=0.99\linewidth, height=1.0\linewidth,
		xmode=log,
		ymode=log,
		xlabel={$\gamma$, regularization parameter},
		ylabel={\# of CG iterations},
		axis y line*=right,
		axis x line*=bottom,
		ylabel near ticks, yticklabel pos=right,
		%ylabel near ticks, ylabel shift={-6pt},
		%legend pos=north east,
		title={Regularization robustness}
		%title={CG iterations to achieve $||H x-b||/||b||<10^{-6}$}
		]
		%\addlegendentry{None}
		\addplot[dash dot, line width=0.5mm] table[x=gamma, y=none] {CGitsvsGamma.dat};
		%\addlegendentry{REG}
		\addplot[dotted, line width=0.5mm] table[x=gamma, y=reg] {CGitsvsGamma.dat};
		%\addlegendentry{PSF (1)}
		\addplot[blue, line width=0.5mm] table[x=gamma, y=psf1] {CGitsvsGamma.dat};
		%\addlegendentry{PSF (5)}
		\addplot[red, line width=0.5mm] table[x=gamma, y=psf5] {CGitsvsGamma.dat};
		%\addlegendentry{PSF (25)}
		\addplot[purple, line width=0.5mm] table[x=gamma, y=psf25] {CGitsvsGamma.dat};
		\end{axis}
		\end{tikzpicture}
		\caption{}
		\label{fig:gamma_sweep}
	\end{subfigure}
	\caption{Left (\ref{fig:cg_convergence}) shows the convergence history for solving a linear system of the form $Hx=b$ using the preconditioned conjugate gradient method, where $b$ has random independent and identically distributed (i.i.d) entries drawn from the standard normal distribution. The y-axis is the relative $l^2$ norm of the residual, i.e., $||b-Hx_k||/||b||$, where $x_k$ is the $k$th iterate. Here the regularization parameter $\gamma$ is determined by the Morozov discrepancy principle. Right (\ref{fig:gamma_sweep}) shows the number of preconditioned conjugate gradient iterations required to reduce the relative $l^2$ norm of the residual below a tolerance of $10^{-6}$, for a range of different regularization parameters $\gamma$. Results in these figures are shown for our preconditioner with 1, 5, and 25 batches of impulse responses (PSF (1), PSF (5), and PSF(25), respectively), regularization preconditioning (REG), and no preconditioning (NONE). The preconditioner is constructed using the Gauss-Newton Hessian. For a fair comparison, the Hessian and Gauss-Newton Hessian are always evaluated at the same parameter $\basalfriction$, which is the optimal point for the inverse problem with $\gamma$ given by the Morozov discrepancy principle.}
\end{figure}

>>>>

\begin{figure}[H]
	\begin{subfigure}{0.5\textwidth}
		\includegraphics[scale=0.23]{meshHeight_view2_edges.png}
		\caption{Bird's eye view}
	\end{subfigure}
	\begin{subfigure}{0.48\textwidth}
		\includegraphics[scale=0.22]{meshHeight_view0_edges_withAxes.png}
		\caption{Top-down view}
	\end{subfigure}
	\caption{The ice sheet, discretized by a mesh of tetrahedra. Color indicates the height of the base of the ice sheet (i.e., the mountain topography). The radius of the domain is $10$ kilometers, and the maximum height of the mountain is $2.1$ kilometers.}
	\label{fig:ice_mountain_mesh}
\end{figure}

@@

The true parameter field, $\basalfriction_\text{true}$, and corresponding velocity field, $\bm{u}_\text{true}$, are shown in Figure \ref{fig:true_beta_u}. We generate synthetic observations by restricting the tangential component of $\bm{u}_\text{true}$ to the top surface to get
\begin{equation*}
\bm{y}_\text{true} = \bm{T} \bm{u}_\text{true} |_{\Gamma_t}.
\end{equation*}
We add $1\%$ multiplicative gaussian noise, $\boldsymbol{\xi}$, to all components of the finite element vector representing $\bm{y}_\text{true}$ to get noisy synthetic observations, $\bm{y}_\text{obs}$, which are used for the inversion. That is, 
\begin{equation*}
\left[\bm{y}_\text{obs}\right]_j := \left[\bm{y}_\text{true}\right]_j + \epsilon \left[\boldsymbol{\xi}\right]_j \left|\left[\bm{y}_\text{true}\right]_j\right|
\end{equation*}
where $\epsilon=0.01$, and the components of the noise function, $\boldsymbol{\xi}_j \sim N(0,1)$, are independent and identically distributed (i.i.d) random numbers drawn from the standard normal distribution. Here $\left[\bm{f}\right]_j$ denotes the $j$th entries of the finite element vector for the discretized versions of the function $\bm{f}$, and $j$ ranges over all entries of the finite element vector. Here we do not use bold to distinguish finite element vectors from their associated functions (as we did previously) because the functions $\bm{y}_\text{true}$, $\bm{y}_\text{obs}$, and $\bm{\xi}$ are already written in bold to denote the fact that they are vector valued.

The reconstructed parameter field $\basalfriction$ is found as the solution to an optimization problem of the following form:
\begin{equation}
	\label{eq:deterministic_optimization_problem}
	\min_{\basalfriction} \quad J(\basalfriction) = J_d(\basalfriction) + J_r(\basalfriction).
\end{equation}
The first term in the objective function is
\begin{equation*}
	J_d(\basalfriction) := \frac{1}{2}\int_{\Gamma_{t}}||\bm{d} - \bm{T}\bm{u}||_{2}^{2}\,dS.
\end{equation*}
We call this term the data misfit, because it measures the mismatch between the observed data and the predicted data based on a candidate parameter $\basalfriction$. Here $\bm{u} = \bm{u}(\basalfriction)$ denotes the velocity field solving $\ref{Stokeseqn}$ for the given parameter $\basalfriction$. The second term,
\begin{equation*}
	J_r(\basalfriction) := \frac{1}{2}\int_{\Omega}\left| \mathcal{L}(\basalfriction - \basalfriction_0)\right|^2\,dx,
\end{equation*}
is a Bilaplacian regularization term. Specifically, $\mathcal{L}$ is the inverse of the solution operator for the following elliptic PDE:
\begin{align*}
-\gamma\Delta u +\delta\,u&=s\,\,\,\,\text{ in }\Gamma_{b},\\
\gamma \bm{\nabla}u\cdot\bm{n}+\mu\,u&=0\,\,\,\text{ on }\partial\Gamma_{b}.
\end{align*}
In all results shown except one, $\gamma$ is chosen so that the solution satisfies the Morozov discrepancy principle, that is,
\begin{equation*}
	\frac{1}{2}\int_{\Gamma_{t}}||\bm{d} - \bm{T}\bm{u}||_{2}^{2}\,dS = \frac{1}{2}\int_{\Gamma_{t}}||\boldsymbol{\xi}||_{2}^{2}\,dS.
\end{equation*}
The value of $\gamma$ satisfying the Morozov discrepancy principle is $\gamma=7.3 \times 10^3$. The exception is Figure \ref{fig:gamma_sweep}, where we vary $\gamma$ to study the effect of changing $\gamma$ on the quality of our preconditioner. The constant $\delta$ is chosen so that the correlation length of functions drawn from $N(0,\mathcal{L}^{-2})$, which is given by the formula $L=\sqrt{\gamma/\delta}$, is $1/10$th of the radius of the domain. The function $\basalfriction_0$ is the constant function with value $\basalfriction_0(x)=10.5$, which is halfway between the maximum and minimum values of $\basalfriction_\text{true}$.

The inverse problem is solved with a preconditioned inexact-Newton conjugate gradient scheme \cite{NOCEDAWRIGHT}. The constant function $\basalfriction_0$ is used as the initial guess. The parameter value is updated iteratively according to the formula
\begin{equation*}
	\basalfriction_{k+1} = \basalfriction_k + \alpha_k p_k,
\end{equation*}
where the search direction $p_k$ is an approximate solution to the Newton linear system
\begin{equation}
	\label{eq:newton_system}
	H p_k = - g_k,
\end{equation}
and $\alpha_k \in \mathbb{R}$ is a step length parameter chosen to satisfy the Armijo conditions \cite{ARMIJO}. Here $g_k$ is the gradient of $J$ with respect to $\basalfriction$ at $\basalfriction_k$, and $H$ is either the Hessian or Gauss-Newton Hessian of $J$ with respect to $\basalfriction$. The Gauss-Newton Hessian is Hessian of $J$, except in the data misfit, $J_d$, the nonlinear mapping $\basalfriction \mapsto \bm{T} \bm{u}$ is replaced by its local linearization at $\basalfriction_k$. The Gauss-Newton Hessian is used for the first five iterations, and the Hessian is used for all subsequent iterations.  The method terminates when $||g_k||/||g_0|| < 10^{-8}$. The Newton systems, \eqref{eq:newton_system}, are solved inexactly using an inner conjugate gradient (CG) iteration, which we terminate when $||g_k + H p_k|| < \min \left(0.5, \sqrt{||g_k||}\right)||g_k||$. The preconditioned conjugate gradient method only requires the application of the (Gauss-Newton) Hessian to vectors. Each application of the (Gauss-Newton) Hessian to a vector requires solving one forward and one adjoint Stokes PDE of the form \eqref{Stokeseqn}, but with different right hand sides.

We use the framework described in this paper to generate Hessian preconditioners. We build H-matrix approximations of the data-misfit Gauss-Newton Hessian (the term in the Gauss-Newton Hessian that arises from the data misfit) using 1, 5, and 25 impulse responses batches. We approximate the Gauss-Newton Hessian rather than the Hessian because the Hessian often has negative values in its integral kernel when one is away from the optimal point. Our results show that our preconditioners, which are built using the Gauss-Newton Hessian, are also good preconditioners for the Hessian. For the purpose of building the Hessian approximation, the basal surface, which is a 2D surface embedded in 3D due to the mountain topography, is flattened into the plane by ignoring the height coordinate. This flattening is necessary because our method requires adding points on the basal surface together to get other points on the basal surface, and this is only well-defined in flat space.

The H-matrix approximation of data misfit Gauss-Newton Hessian is added to a H-matrix approximation of the Hessian of the regularization term, $J_r$, to form a H-matrix approximation of the Gauss-Newton Hessian. For the Hessian of the regularization term, straightforward calculation shows that the discrete regularization term in both the Hessian and the Gauss-Newton Hessian is given by
$
(\gamma \bm{K} + \delta \bm{M}) \bm{M}^{-1} (\gamma \bm{K} + \delta \bm{M})
$
where $\bm{K}$ is a finite element stiffness matrix with appropriate boundary conditions, and $\bm{M}$ is the finite element mass matrix. These matrices are sparse, so we form H-matrix approximations of these matrices using standard techniques for approximating sparse matrices with H-matrices. Then we invert $\bm{M}^{-1}$, and multiply these matrices together using H-matrix methods. We factor the overall Gauss-Newton Hessian approximation using fast H-matrix methods, then and use the factorization as a preconditioner \cite{HLIBPRO}. The resulting preconditioners for 1, 5, and 25 impulse response batches are denoted by PSF (1), PSF (5), and PSF (25), respectively.

In Table \ref{tab:newton_convergence_table} we study the performance of our preconditioner for accelerating the solution of optimization problem \eqref{eq:deterministic_optimization_problem}. We build the PSF (5) preconditioner in the third Gauss-Newton iteration, and re-use it for all subsequent Gauss-Newton and Newton iterations. No preconditioner is used in the three Gauss-Newton iterations before the PSF (5) preconditioner is built. We compare our method with the most commonly used existing preconditioners, which are no preconditioning (denoted by NONE), and preconditioning by the regularization term in the Hessian (denoted REG). In Figure \ref{fig:newton_iterates} we show several Newton iterates using our PSF (5) preconditioner. Our results show that our PSF (5) preconditioner substantially outperforms the no preconditioning and regularization preconditioning. Using PSF (5) reduces the total number of Stokes PDE solves by a factor of six compared to no preconditioning, and by a factor of nine compared to regularization preconditioning. Interestingly, we observe that no preconditioning actually outperforms regularization preconditioning here because the noise level, and hence the regularization parameter, $\gamma$, are small.

Next, we build PSF (1), PSF (5), and PSF (25) preconditioners based on the Gauss-Newton Hessian evaluated at the converged solution $\basalfriction$. We use the preconditioned conjugate gradient method to solve a linear system with the Hessian as the coefficient operator and a right hand random vector with i.i.d. entries drawn from the standard normal distribution. In Figure \ref{fig:cg_convergence} we compare the convergence of conjugate gradient for solving this linear system using the PSF (1), PSF (5), PSF (25), REG, and NONE preconditioners. The preconditioned conjugate gradient method converges fastest with the PSF preconditioners, with PSF (25) converging fastest, followed by PSF (5), followed by PSF (1), as expected. Preconditioned conjugate gradient converges much slower with no preconditioning and regularization preconditioning than it does with PSF preconditioning, with no preconditioning outperforming regularization preconditioning. In Figure \ref{fig:gamma_sweep}, we perform preconditioned conjugate gradient solves on linear systems of the same form, except now we vary $\gamma$. The performance of our PSF preconditioners is good and relatively stable over a wide range of $\gamma$'s. All PSF preconditioners perform the same for medium and large values of $\gamma$. For small values of $\gamma$ PSF (25) performs slightly better than PSF (5), which performs slightly better than PSF (1). As expected, regularization preconditioning performs well for large $\gamma$ and poorly for small $\gamma$. Our PSF preconditioners outperform no preconditioning and regularization preconditioning for medium and small $\gamma$, and perform similarly to regularization preconditioning for large $\gamma$.

\begin{figure}[H]
	\begin{subfigure}{0.24\textwidth}
		\includegraphics[scale=0.1]{m1_psf.png}
		%		\caption{}
		%		\label{fig:newtoniter1}
	\end{subfigure}
	\begin{subfigure}{0.24\textwidth}
		\includegraphics[scale=0.1]{m3_psf.png}
		%		\caption{}
		%		\label{fig:newtoniter2}
	\end{subfigure}
	\begin{subfigure}{0.24\textwidth}
		\includegraphics[scale=0.1]{m4_psf.png}
		%		\caption{}
		%		\label{fig:newtoniter3}
	\end{subfigure}
	\begin{subfigure}{0.24\textwidth}
		\includegraphics[scale=0.1]{m6_psf.png}
		%		\caption{}
		%		\label{fig:newtoniter4}
	\end{subfigure}
	%\begin{subfigure}{0.05\textwidth}
	%	\includegraphics[scale=0.1]{mColorBar.png}
	%\end{subfigure}
	\caption{The log basal friction parameter $\basalfriction$ after $1$ (left), $3$ (second from left), and $4$ (third from left) Newton iterations, and the final Newton iterate (right). The PSF (5) preconditioner is constructed between Newton iterations 3 and 4 and re-used for all subsequent iterations.}
	\label{fig:newton_iterates} 
\end{figure}

\begin{table}[H]
	\begin{center}
		\begingroup
		\setlength{\tabcolsep}{3pt}
		\renewcommand{\arraystretch}{1.25}
		\begin{tabular}{c| c c c | c c c | c c c}
			&  \multicolumn{3}{|c|}{PSF (5)} & \multicolumn{3}{|c|}{REG} & \multicolumn{3}{|c}{NONE} \\
			\hline
			Iter & 
			\#CG & \#Stokes & $||g||$ & 
			\#CG & \#Stokes & $||g||$ & 
			\#CG & \#Stokes & $||g||$ \\
			\verb~0~ &
			\verb~1~ & \verb~4~ & \verb~2.0e+7~ &
			\verb~1~ & \verb~4~ & \verb~2.0e+7~ &
			\verb~1~ & \verb~4~ & \verb~2.0e+7~ \\
			\verb~1~ &
			\verb~2~ & \verb~6~ & \verb~5.8e+6~ &
			\verb~2~ & \verb~6~ & \verb~7.9e+6~ &
			\verb~2~ & \verb~6~ & \verb~5.8e+6~ \\
			\verb~2~ &
			\verb~4~ & \verb~10~ & \verb~2.4e+6~ &
			\verb~5~ & \verb~12~ & \verb~3.9e+6~ &
			\verb~4~ & \verb~10~ & \verb~2.4e+6~ \\
			\verb~3~ &
			\verb~2~ & \verb~6+22~ & \verb~5.8e+5~ &
			\verb~13~ & \verb~28~ & \verb~1.6e+6~ &
			\verb~13~ & \verb~28~ & \verb~5.8e+5~ \\
			\verb~4~ &
			\verb~5~ & \verb~12~ & \verb~5.6e+4~ &
			\verb~42~ & \verb~86~ & \verb~4.8e+5~ &
			\verb~33~ & \verb~68~ & \verb~9.1e+4~ \\
			\verb~5~ &
			\verb~10~ & \verb~22~ & \verb~3.5e+3~ &
			\verb~84~ & \verb~170~ & \verb~7.7e+4~ &
			\verb~57~ & \verb~116~ & \verb~6.2e+3~ \\
			\verb~6~ &
			\verb~14~ & \verb~30~ & \verb~2.7e+1~ &
			\verb~125~ & \verb~252~ & \verb~5.0e+3~ &
			\verb~76~ & \verb~154~ & \verb~1.1e+2~ \\
			\verb~7~ &
			\verb~0~ & \verb~2~ & \verb~4.1e-2~ &
			\verb~194~ & \verb~390~ & \verb~7.9e+1~ &
			\verb~117~ & \verb~236~ & \verb~2.3e-1~ \\
			\verb~8~ & 
			\verb~---~ & \verb~---~ & \verb~---~ &
			\verb~0~ & \verb~2~ & \verb~1.6e-1~ &
			\verb~0~ & \verb~2~ & \verb~4.1e-2~ \\
			\hline
			Total & 
			\verb~38~ & \verb~114~ & \verb~---~ &
			\verb~466~ & \verb~950~ & \verb~---~ &
			\verb~303~ & \verb~624~ & \verb~---~ \\
		\end{tabular}
		\endgroup
	\end{center}
	\caption{Convergence history for solving the Stokes inverse problem using the preconditioned inexact Newton-Krylov method. The Newton iteration terminates when the norm of the gradient has decreased from its initial value by a factor of $10^{-8}$. The Gauss-Newton Hessian is used in iterations $0$--$5$, and the Hessian is used in all subsequent iterations. Preconditioners shown are our method with five impulse response batches (PSF (5)), regularization preconditioning (REG), and no preconditioning (NONE). Our preconditioner (PSF (5)) is built at iteration $3$ using the Gauss-Newton Hessian, and re-used for all subsequent iterations. The iterates are $\basalfriction_0$ (initial guess), $\basalfriction_1$ (parameter after one Newton step), $\basalfriction_2$ (parameter after two Newton steps), and so on. The columns titled \#CG show the number of conjugate gradient iterations used to solve the Newton system at $\basalfriction_k$, for the purpose of computing a search direction towards $\basalfriction_{k+1}$. The columns titled $||g||$ show the $l^2$ norm of the gradient at $\basalfriction_k$. The columns titled \#Stokes show the number of Stokes PDE solves performed in each Newton iteration. This consists of Stokes solves for $u$ performed during the linesearch going from $\basalfriction_{k-1}$ to $\basalfriction_k$ (including one Stokes solve at $\basalfriction_k$), plus one adjoint Stokes solve to compute the gradient at $\basalfriction_k$, plus one incremental forward and one incremental adjoint Stokes solve per CG iteration for solving the Newton system at iteration $k$. In the PSF (5) portion of row 3, we write $6+22$ to indicate that 6 Stokes solves were used during the standard course of the iteration, and 22 additional Stokes solves were used to build the PSF (5) preconditioner. }
	\label{tab:newton_convergence_table}
\end{table}
